Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-13 paper-ieee-trans.tex
cris comments

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e-Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission-funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The base of knowledge collected
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, motivated
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

@@

Soon it become apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of teams
and deployment of pilot testbeds.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed and aware to software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities, including physics, biology and
earth sciences, with intensive computing and large-scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address these requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

@@

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, thus not
considered by the project, and as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nevertheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to
achieve a higher maturity level \cite{cmm}.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release.
This procedure aimed to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, to validate and distribute the software to
the production infrastructure. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this paper. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, will provide guidelines to develop quality
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

%The {\sl INDIGO--DataCloud} project reflects the progress made in software
%quality and reliability aspects throughout the past scientific European
%experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
%highly influenced by the insights of current big worldwide collaborations of
%software development. These collaborations have continuously inspired the use of new
%technologies and procedures to increase the robustness and quality of the
%software being delivered. In this scenario, the implementation of the DevOps culture
%is one of the major outcomes, as this paradigm has been progressively adopted by the
%project.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's
behaviour throughout the development, release and post--release stages.

@@

The SQA criteria is a set of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the produced software components.
It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, the criteria promotes the adoption of the software
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment.

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the successful execution of the previous
functionality tests already present in the code to uncover any malfunction introduced by
the changes. \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment,
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria requires the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\textit{Development metrics} are obtained programatically from several sources, namely GitHub
API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
completeness, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

\textit{Release metric} sources are the online repository servers, namely the
Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
server also contains valuable release data as it is the service where the packages
and containers are being built before being uploaded to the online repositories. Release
KPIs primarily focus on the frequency and efficiency -- mainly rollbacks -- statistics.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud}’s application repository.

@@

The released software is tested in production environments through the
Staged--Rollout process, before inclusion in the EGI's Cloud Middleware Distribution (CMD).
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.

@@

{\sl INDIGO--DataCloud} project has received funding from the European Research
Council (ERC) under the European Union's Horizon 2020 research and innovation
programme (grant agreement n° 777536).

@@

\bibitem{globus}
I. Foster and C. Kesselman, \emph{Globus: a Metacomputing Infrastructure
Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2,
pp. 115–128, 1997.

@@

\bibitem{glite}
E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A.,
Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1),
pp. 33–45, Scientific Publishers OWN, 2006.

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70,
Eds. John Wiley \& Sons Inc., 2002.

@@

\bibitem{etics}
A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
the international software engineering service for the Grid}, in Journal of
Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.

@@

\bibitem{mario}
M. David et al, \emph{Validation of Grid Middleware for the European Grid
Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543–558,
Springer, 2014.

@@

\bibitem{disvis} G.C.P. Van Zundert, A.M.J.J. Bonvin,
DisVis: quantifying and visualizing the accessible interaction space of distance
restrained biomolecular complexes, Bioinformatics 31 (2015) 3222–3224

\bibitem{powerfit} G.C.P. Van Zundert, A.M.J.J. Bonvin,
Fast and sensitive rigid-body fitting into cryo-EM density maps with PowerFit,
AIMS Biophys. 2 (2015) 73–87.

>>>>

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e-Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission-funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the European Union's funded 5th, 6th and 7th Framework Programmes,
and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

@@

Soon it became apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the improvement of the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of development teams
and deployment of pilot testbeds.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities, including physics, biology and
earth sciences, with intensive computing and large-scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

@@

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
considered by the project, and as such the project's
development design suffered from the lack of the methods, tools, techniques and best
practices emerging from these new discipline of software engineering
\cite{agile}. Nevertheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to
achieve a higher maturity level \cite{cmm}.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements coming from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, which issued more than 10 thousand bugs fixes, 1.7 thousand patches and
defined over 300 development tasks tracked using bug/task management tools.

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release.
This procedure aimed to improve the reliability of the software components by applying
acceptance criteria checks at each stage of the software development lifecycle
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would had been accomplished through the
development and maintenance of some operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this paper. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures administrators and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
quality of the solutions implemented. All the above with the additional complexity of being framed into a
large collaborative project with an extensive list of partners.

In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, will provide guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

%The {\sl INDIGO--DataCloud} project reflects the progress made in software
%quality and reliability aspects throughout the past scientific European
%experiences. Nevertheless (I would remove this), the project???s Quality Assurance procedures are also
%highly influenced by the insights of current big worldwide collaborations of
%software development. These collaborations have continuously inspired the use of new
%technologies and procedures to increase the robustness and quality of the
%software being delivered. In this scenario, the implementation of the DevOps culture
%is one of the major outcomes, as this paradigm has been progressively adopted by the
%project.

@@

The initial software quality policies were defined in the first deliverable document of the
project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's
behaviour throughout the development, release and post--release stages.

@@

The SQA criteria is a set of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the produced software components.
It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, there are provided the minimum requirements on documentation content and it is stresed the importance of the usage of automated
solutions for software deployment.

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition, and periodic execution, of tests for the fixed issues that ensure these issues are not reintroduced during the development activities.
 \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment,
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also 
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria requires the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\textit{Development metrics} are obtained programatically from several sources, namely GitHub
API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
fulfilment, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

\textit{Release metric} sources are the online repository servers, namely the
Linux packages \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
server also contains valuable release data as it is the service where the packages
and containers are being built before being uploaded to the online repositories. Release
KPIs primarily focus on the frequency and efficiency -- mainly rollbacks -- statistics.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud}???s application repository.

@@

The released software is tested in production environments through the
Staged--Rollout process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.

@@

{\sl INDIGO--DataCloud} project has received funding from the European Research
Council (ERC) under the European Union's Horizon 2020 research and innovation
programme (grant agreement n?? 777536).

@@

\bibitem{globus}
I. Foster and C. Kesselman, \emph{Globus: a Metacomputing Infrastructure
Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2,
pp. 115???128, 1997.

@@

\bibitem{glite}
E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A.,
Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1),
pp. 33???45, Scientific Publishers OWN, 2006.

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63???70,
Eds. John Wiley \& Sons Inc., 2002.

@@

\bibitem{etics}
A. Di Meglio, M.-E.B??gin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
the international software engineering service for the Grid}, in Journal of
Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.

@@

\bibitem{mario}
M. David et al, \emph{Validation of Grid Middleware for the European Grid
Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543???558,
Springer, 2014.

@@

\bibitem{disvis} G.C.P. Van Zundert, A.M.J.J. Bonvin,
DisVis: quantifying and visualizing the accessible interaction space of distance
restrained biomolecular complexes, Bioinformatics 31 (2015) 3222???3224

\bibitem{powerfit} G.C.P. Van Zundert, A.M.J.J. Bonvin,
Fast and sensitive rigid-body fitting into cryo-EM density maps with PowerFit,
AIMS Biophys. 2 (2015) 73???87.
